$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-09-04 paper/paper.tex
First go at an improved example

\begin{example}[Typing derivation]
  \label{example-derivations}
  In \autoref{lst:derivation} we illustrate a typing derivation using shared, graded and linear types.
  We use $\func{\omega}$ as a shortcut for $\one \comma \one$ in an algebra of shared types.

  \begin{lstlisting}[label=lst:derivation,mathescape,numbers=left,caption=Typing two processes that send parts of one channel over another.]
  $\func{sync} \; : \; \forall \{n\} \to \Var_n \to \Var_n \to \Var_n \to \Process_n$
  $\func{sync} \; i0 \; i1 \; o \; =$
    $\recv{i0}{} \recv{(\suc \; i1)}{} \send{(\suc \suc \; o)}{\suc \constr{0}} \send{(\suc \suc \; o)}{\constr{0}} \PO$

  $\func{send} \; : \; \forall \{n\} \to \Var_n \to \Process_n$
  $\func{send} \; c \; = \new{} (\send{\suc c}{\constr{0}} \PO)$

  $\func{recv} \; : \; \forall \{n\} \to \Var_n \to \Process_n$
  $\func{recv} \; c \; = \recv{c}{} (\recv{\suc c} \PO)$

  $\func{example} \; : \; \Process_{\constr{0}}$
  $\func{example} \; = \new{} ( \func{send} \; \constr{0}$
           $\comp{}{ \new{} ( \func{send} \; \constr{0} }$
             $\comp{}{ \new{} ( \func{recv} \; \constr{0} }$
               $\comp{}{ \func{sync} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}))) }$

  $\func{\_} \; : \; \types{\constr{[]}}{\constr{[]}}{\func{example}}{\constr{[]}}$
  $\func{\_} = \new{} \; \channel{\unit}{\omega} \; \lz \; 1$
    $( (\send{\suc \constr{0}}{\constr{0}} \PO)$
    $\comp{}{ \new{} \; \channel{\unit}{\omega} \; \lz \; 1 }$
      $( (\send{\suc \constr{0}}{\constr{0}} \PO)$
      $\comp{}{ \new{} \; \channel{\unit}{\omega} \; \lz \; 2 }$
        $( (\recv{\constr{0}}{} (\recv{\suc \constr{0}}{} \PO))$
        $\comp{}{( ( \recv{\suc (\suc \constr{0}))}{}}$
          $( \recv{\suc (\suc \constr{0}))}{}$
          $( \send{\suc (\suc \constr{0}))}{\suc \constr{0}}$
          $( \send{\suc (\suc \constr{0}))}{\constr{0}} \PO))))$
  \end{lstlisting}

  Notice that the \emph{only} differences between the definition of the untyped process $\func{P}$ and its typing derivation are the extra type annotations that accompany scope restriction.
  This extra information demands a (limited) choice.
  To type $\func{P}$, we need a channel over which we can transmit another channel.
  We therefore create a new channel with usage annotations $\one \comma \one$ capable of transmitting data of type $\channel{\unit}{\func{\omega}}$, and declare (with $\li$) that we only transmit input capabilities over it (line 9).
  The left-hand process (line 3) is typed following its syntax (line 10).
  For the right-hand (lines 4-5) we need the channel that is going to be sent (lines 11-12).
  The type $\unit$ of this channel is entirely determined by the type $\channel{\unit}{\func{\omega}}$ of the channel over which it is going to be transmitted.
  Its usage annotation is $\one \comma \one$: we will transmit $\one \comma \zero$ and save $\zero \comma \one$ for the continuation process.

>>>>

\begin{example}[Typing derivation]
\label{example-derivations}

We present an example with four parallel processes: two processes that send data over a channel; one that receives data over a channel twice; and a fourth process that sticks the previous processes together and synchronizes them.

We start by defining the process doing the synchronization.
This process sequentially receives data on both input channels and then sends on the output channel both pieces of received data.
For simplicity, when we type this process we instantiate it to variable references and usage annotations.
We expect both input channels to be graded and capable of performing input once.
The output channel must be capable of performing output twice.
Notice that the well-typedness proof has no choice but to blindly follow the syntax.

\begin{flalign*}
& \func{sync} \; : \; \Var_n && \to \Var_n \to \Var_n \to \Process_n & \\
& \func{sync} \; in0 \; in1 \; out \; &&= \\
& && \recv{in0}{} \\
& && \recv{(\suc \; in1)}{} \\
& && \send{(\suc \suc \; out)}{\suc \constr{0}} \\
& && \send{(\suc \suc \; out)}{\constr{0}} \PO \\
\end{flalign*}
\begin{flalign*}
& \func{\vdash sync} && : \types{\gamma \comma \channel{t}{\lz} \comma \channel{t}{\lz} \comma \channel{t}{\lz}\\ & &&}{\Gamma \comma (1 \comma 0) \comma (1 \comma 0) \comma (0 \comma 2)\\ & &&}{\func{sync} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}\\ & &&}{\Gamma \comma (0 \comma 0) \comma (0 \comma 0) \comma (0 \comma 0)} & \\
& \func{\vdash sync} && = \\
& && \recv{\suc \suc \constr{0}}{} \\
& && \recv{\suc \suc \constr{0}}{} \\
& && \send{\suc \suc \constr{0}}{\suc \constr{0}} \\
& && \send{\suc \suc \constr{0}}{\constr{0}} \PO \\
\end{flalign*}

Next, we define the processes doing the input and output.
This time too, we instantiate them to concrete variable references for typing.

\begin{flalign*}
& \func{recv} \; : \; \Var_n \to \Process_n & \\
& \func{recv} \; c \; = \recv{c}{} (\recv{\suc c} \PO) \\
& \\
& \func{\vdash recv} \; \types{\gamma \comma \channel{t}{\lz}}{\Gamma \comma (2 , 2)}{\func{recv} \; \constr{0}}{\Gamma \comma (0 \comma 2)} & \\
& \func{\vdash recv} \; = \; \recv{\constr{0}}{} \recv{(\suc \constr{0})}{} \PO \\
& \\
& \func{send} \; : \; \Var_n \to \Var_n \to \Process_n & \\
& \func{send} \; c \; v \; = \send{c}{v} \PO \\
& \\
& \func{\vdash send} \; \types{\gamma \comma \channel{t}{\lz} \comma t}{\Gamma \comma (1 , 1) \comma \lz}{\func{send} \; (\suc \constr{0}) \; \constr{0}}{\Gamma \comma (1 \comma 0)\comma \lz} & \\
& \func{\vdash send} \; = \; \send{(\suc \constr{0})}{\constr{0}} \PO
\end{flalign*}

Finally, we put the four processes together in parallel.
It is here that the well-typedness proof must make a choice as to what types and usage annotations to instantiate new channels with.
In this case, these types and usage annotations were determined by the types of the previous subproofs, and we can thus omit most of them with $\_$ and have Agda find them.
We use $\func{\omega}$ as a shortcut for $\one \comma \one$ in an algebra of shared types.

\begin{flalign*}
& \func{example} && : \; \Process_{\constr{0}} & \\
& \func{example} && = \new{} ( ( \new{} (\func{send} \; \suc \constr{0} \; \constr{0})) \\
& && \comp{}{ \new{} ( (\new{} (\func{send} \; \suc \constr{0} \; \constr{0} })) \\
& && \comp{}{ \new{} ( \func{recv} \; \constr{0} } \\
& && \comp{}{ \func{sync} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}))) } \\
\end{flalign*}
\begin{flalign*}
& \func{\vdash example} && : \; \types{\constr{[]}}{\constr{[]}}{\func{example}}{\constr{[]}} & \\
& \func{\vdash example} && = \new{} \; \_ \; \_ \; \_ \; (( \new{} \; \_ \; \_ \; \_ \; \func{\vdash send}) \\
& && \comp{}{ \new{} \; \_ \; \_ \; \_ } \; (( \new{} \; \_ \; \_ \; \_ \; \func{\vdash send}) \\
& && \comp{}{ \new{} \; \channel{\unit}{\func{\omega}} \; \_ \; \_ } \; ( \func{\vdash recv} \\
& && \comp{}{ \func{\vdash sync} }))) \\
\end{flalign*}
